---
output:
  pdf_document:
    toc: true
    toc_dept: 3
    number_sections: true
    
fontsize: 12pt
mainfont: Times New Roman
linestretch: 1.5
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```
```{r, include=FALSE}
# Library and Import Dataset
library(readxl)
library(knitr)
library(tidyverse)
library(kableExtra)
library(ggplot2)
library(e1071)
library(dplyr)
library(ggpubr)
library(car)
library(explore)
library(DataExplorer)
df <- read_xlsx('~/R/Group/Sleep_Efficiency.xlsx')
```
\newpage
# Introduction

## Research Background

Plenty of research regarding sleep has been conducted throughout the decades. Typically, the research falls within the context of the importance of sleep, the factors affecting sleep and the cause & effect of sleep deprivation. Based on Naito et al. (2021) research paper, the paper highlights the basic statistical aspect of sleep deprivation among the undergraduate students in Malaysia with a sample size of 1132 participants. The paper returned stating that 58.1% of the participants responded with sleep deprivation. Furthermore, the paper found that the people living within the Asian region typically sleep 6 hours 30 minutes. On average, a difference of 1 hours 30 minute sleep deprivation (Recommended Sleep: 8 hours). The repercussions typically affect the individual person's cognitive performance, emotional state, reaction speed and excessive daytime sleepiness (Cardoso 2018; Costa et al. 2022). Thereby suggesting a significant deterioration of the individual well being and overall performance. To emphasize the ramification of sleep deprivation, Ellenbogen (2005) paper highlights that the sample made 36% more error when sleep deprived.

## Literature Review

The intricate relationship between sleep quality and its quantifiable metrics has been a focal point of contemporary sleep research. Initial studies, such as those by Naito et al. (2021), posited a direct correlation between sleep duration—specifically, 7-9 hours per night—and optimal cognitive performance during waking hours. However, this perspective has been challenged by further research , which revealed that an increase in sleep duration does not necessarily equate to improved sleep quality or cognitive function (Alfonsi et al., 2021). This conflict underscores the complexity of sleep quality as a multidimensional construct, extending beyond mere sleep duration.
\newline

Therefore, the concept of sleep efficiency emerged as an important measure that developed a more detailed approach of quantification of sleep quality according to duration and disturbances. The comparison study on varying contexts, such as the pandemic (Alfonsi et al., 2021) and periods of differing emotional states (Okamura et al., 2020; Hofer and Chen, 2020), highlighted the influence of external factors on sleep efficiency. Sleep latency, of which stress and happiness work on as a factor, has come to be recognized as a factor of significance that indicates sleep quality is not known only by duration of sleep.
\newline

Technological advancements in sleep measurement, as evidenced by Meltzer et al. (2005), have facilitated a more detailed examination of sleep patterns and quality. Despite the discrepancies in accuracy between commercial polysomnography (PSG) devices and consumer-grade wearables, the evolution of such technologies indicates a promising avenue for non-invasive sleep monitoring. However, current literature reveals a notable gap in understanding the relationship between REM sleep, sleep stages, and sleep efficiency on sleep quality. Previous research has primarily focused on indirect aspects of REM sleep or specific demographics without a broader exploration of these variables. By delving into these under explored areas, we aim to contribute to a more holistic understanding of sleep quality, ultimately developing a model that resembles the features of a non-invasive arm sensor (ACTG) on an Apple watch.

## Problem Statement

1. From a user-centric perspective, the current measurement metric to evaluate sleep quality does not provide any added value to users. Thus, the project aims to research an alternate measurement metric to quantify sleep quality which is sleep efficiency.

## Research Objective

1. Explore and identify which factors have significant influences on sleep efficiency. 
\newline
2. Explore if there is any positive correlation between total hours sleep and sleep efficiency as both are a measurement of sleep quality. 
\newline
3. Explore and identify the optimize model upon conducting the iteration process. 
\newline
4. Implement a model with a similar feature with apple watch which emulates a non-invasive arm sensor (ACTG) and evaluate the model.

## Research Question
The primary question guiding this question is: How does the total amount of sleep influences sleep efficiency? The investigation will delve into conducting an exploratory data analysis (EDA) followed by a statistical modeling to identify the significant impact from a statistical perspective. Moreover, it will extend into predicting sleep efficiency and analyzing the results based on the collected data.

## Data Overview

The dataset currently has 15 variables and 452 observations. Table 1 provides the description of each variable.
```{r}
#Table 1: Variable description
No <- 1:15


Var <- c('ID',
              'Age',
              'Gender',
              'Bedtime',
              'Wakeup Time',
              'Sleep Duration',
              'Sleep Efficiency',
              'REM Sleep Percentage',
              'Deep Sleep Percentage',
              'Light Sleep Percentage',
              'Awakenings',
              'Caffeine consumption',
              'Alcohol consumption',
              'Smoking status',
              'Exercise frequency'
              )


Des <- c('A unique identifier for each test subject',
                 'Age of the test subject','Male or Female (0 = Female, 1 = Male)',
                 'The time the test subject goes to bed each night',
                 'The time the test subject wakes up each morning',
                 'The total amount of time the test subject slept (in hours)',
                 'A measure of the proportion of time in bed spent asleep',
                 'The percentage of total sleep time spent in REM sleep',
                 'The percentage of total sleep time spent in deep sleep',
                 'The percentage of total sleep time spent in light sleep',
                 'The number of times the test subject wakes up during the night',
                 'The amount of caffeine consumed in the 24 hours prior to bedtime (in mg)',
                 'The amount of alcohol consumed in the 24 hours prior to bedtime (in oz)',
                 'whether or not the test subject smokes (0 = No, 1 = Yes)',
                 'The number of times the test subject exercises each week'
                 )


dd.df<- data.frame(No,Var,Des)


kable(dd.df,
      align='lll',
      col.names=c('No','Variable','Description'),
      caption = 'Data Description')
```

## Exploratory Data Analysis
### Sleep Efficiency & Deep sleep percentage

Based on the scatter plot, it can be observed that two clusters have formed indicating that deep sleep percentages have a significant relation with sleep efficiency. It can be inferred that users with higher percentages of deep sleep have a positive correlation with sleep efficiency.

```{r,echo=FALSE, message=FALSE, warning=FALSE}
#Image 1: Scatter plot (SEFF vs Deep sleep %)
df %>%
  ggplot(aes(y=`Sleep efficiency`,
             x=`Deep sleep percentage`,
             color = 'lightblue'))+
  geom_point()+
  theme_update()+
  labs(caption = 'Image 1: Scatter Plot (Sleep Efficiency in relation to Deep Sleep %)', y = 'Sleep Efficiency', x = 'Deep Sleep Percentage')+
  theme(plot.caption = element_text(hjust = 0))+
  guides(color = FALSE)
```

### Awakenings vs Sleep Efficiency
Based on the boxplot, participants with the least number of awakening within their sleep had the higher sleep efficiency percentage.

```{r}
#Image 2
df %>%
  na.omit() %>%
  ggplot(aes(factor(Awakenings),`Sleep efficiency`, fill = as.factor(Awakenings)))+
  geom_boxplot()+
  theme_update()+
  labs(caption = 'Image 2: Boxplot (Sleep Efficiency in relation to Awakening)',
       y = 'Sleep Efficiency %',
       x = 'Number of awakening')+
  theme(plot.caption = element_text(hjust = 0))

```

### Correlation Analysis

The correlation matrix illustrates the pairwise correlation coefficients between various variables. The insights are separated according to the strength of the correlations between the independent variables and the dependent variable, sleep efficiency.Based on the correlation analysis, it can be seen that Deep Sleep Percentage, Light Sleep Percentage, Awakening shows significant correlation with the dependent variable. REM Sleep percentage and Sleep Duration does not hold any significant correlation however the literature review supported the correlation among the variables therefore the model would continue with the current response variables.

```{r, echo=FALSE, message=FALSE}
#Image 3
plot_correlation(na.omit(df), type = "c")
```

# Regression Analysis
## Model Approach

Regression analysis is a statistical method for identifying and quantifying the relationship between the dependent and independent variables. However, the variable can be categorized into two data structures which are categorical and numerical data (Sharda et al. 2017). Therefore, the analysis consists of two sub analysis techniques for the respective data structure which are known as Logistic Regression and Linear Regression. Within the context of the project, linear regression would be the focus as 73% of the data are continuous (Numeric) and the remaining 27% are discrete data (Categorical). Linear regression assumes a linear relationship between the independent and dependent variables within the model thereby expliciting stating that it is only suitable for continuous data type. Within the medical field, data are typically recorded on a numeric data type, therefore studies such as Hockenberry et al (2012) and Lee (2022) conducted the linear regression as the main method to analyze and quantify the relationship. 

## Variable Selection

Based on literature review conducted during the proposal, the project objective would be to conduct the research based on the response variables. The response variables would serve as the initial models. The objective of pre-selecting the variable serves to understand the significant differences between the dependent and independent variable based on literature review thereby the remaining independent variables would be excluded. Based on Lee (2022) studies, variable selection can be conducted in two manners which are forward and backward selection. Forward selection would be adding a significant independent variable and the process is done in iteration until the independent variable is fully incorporated into the response variable. Backward selection operates similarly to forward selection; however, instead of adding, it removes the independent variable based on the least significant differences. The current project would be implementing the backward selection method to remove the independent variable in search for the best optimize model.
 
```{r}
#Table 2: Response Variables
No_rv <-1:6


R_var <- c('Sleep Efficiency',
           'REM sleep percentage',
           'Sleep Duration',
           'Deep sleep percentage',
           'Light Sleep Percenatge',
           'Awakening')

type <- c('Continuous',
          'Continuous',
          'Continuous',
          'Continuous',
          'Continuous',
          'Discrete')

rv.df <- data.frame(No_rv,R_var, type)


kable(rv.df,
      align='llll',
      col.names = c('No','Response Variable','Type'),
      caption = 'Selection of Response Variable')
```

## Model Output
```{r, include=FALSE}
#Data Problem
df_clean <- na.omit(df)
df_clean <- unique(df_clean)


#Removal of unwanted column
df_clean$ID = NULL
df_clean$`Wakeup time`=NULL
df_clean$Bedtime = NULL


#Convert variable in factors
df_clean$Gender <- factor(df_clean$Gender)
df_clean$Awakenings <- factor(df_clean$Awakenings)
df_clean$`Alcohol consumption` <- factor(df_clean$`Alcohol consumption`)
df_clean$`Caffeine consumption`<- factor(df_clean$`Caffeine consumption`)
df_clean$`Exercise frequency` <- factor(df_clean$`Exercise frequency`)
```
```{r, include=FALSE}
df_res <- df_clean %>%
  select(`Sleep efficiency`,
         `REM sleep percentage`,
         `Sleep duration`,
         `Deep sleep percentage`,
         `Light sleep percentage`,
         Awakenings)

# Model 3
best.model <- lm(df_res$`Sleep efficiency` ~`REM sleep percentage`+`Deep sleep percentage`+Awakenings, df_res)
summary(best.model)

```

```{r}
#Table 9: Coefficient table of Model 3'
no_bm <- 1:9


var_bm <- c('Intercept',
            'REM sleep %',
            'Deep sleep %',
            'Awakenings1',
            'Awakenings2',
            'Awakenings3',
            'Awakenings4',
            'Adjusted R-squared',
            'p-value')


est_bm <- c('0.3861391',
            '0.0070839',
            '0.0059850',
            '-0.0563633',
            '-0.1314648',
            '-0.1317450',
            '-0.1242474',
            '-',
            '-')

pr_bm <- c('< 2e-16',
           '6.74e-13',
           '< 2e-16',
           '2.33e-10',
           '< 2e-16',
           '< 2e-16',
           '< 2e-16',
           '-',
           '-')

r2 <- c('-',
        '-',
        '-',
        '-',
        '-',
        '-',
        '-',
        '0.7867',
        '-')

p_value1 <- c('-',
              '-',
              '-',
              '-',
              '-',
              '-',
              '-',
              '-',
              '< 2.2e-16')


df_bm<- data.frame(no_bm,var_bm,est_bm,pr_bm, r2, p_value1)


kable(df_bm,
      align='lll',
      col.names=c('No','Variable','Estimate','Pr(>|t|)', 'Adjust R Square', 'p-value'),
      caption = 'Coefficient table of Model 3')
```

## Model Assumption
**1. Normality - QQ Plot**
\newline
Based on the Q-Q plot, it can be inferred that majority of the data points lies within normality.

```{r}
#Image 4 - QQplot
ggqqplot(best.model$residuals, caption = 'Image 4: QQ Plot of Residue')+
  labs(x = 'Theoretical Quantile', y = 'Data Quantile')+
  theme_update()+
  theme(plot.caption = element_text(hjust = 0))
```
\newpage

**2. Linearity** 
\newline
Based on the Added-Variable Plot (AVP), REM Sleep percentage reflects the ideal scatter plot for linearity. The other independent variable reflects two or more clusters within the AVP, although it is not ideal, the model was still able to detect and draw the best fitted line to determine the relationship in a linearity aspect. 

```{r, warning=FALSE}
#Image 5
avPlots(best.model)
```

\newpage

**3. Homoscedasticity**
\newline
Based on the Residual vs Fitted plot (RVF), the scatter plot does not adhere to a standard RVP which ideally is a consistent spread of residuals across all levels of predicted values which typically represents homoscedasticity. On the other hand, the changing spread of residuals as predicted values change as the x-axis extends. It would represent heteroscedasticity, inferring that the model error increases significantly as the boundary of the residual increases. Within this scenario, the model can pass as homoscedasticity as the boundary of the residue remains consistent.

```{r}
#Image 6
plot(best.model, which = 1)
```

**4. Independence**
\newline
To evaluate the independence of the model, the durbin watson test is used to evaluate the autocorrelation within the model. Based on the result, the D-W statistic achieves a value of 2.09309 (very near to 2), an autocorrelation of -0.0474847 and a p-value of 0.344. It can be concluded that the model does not inherit any autocorrelation because the D-W statistic is near 2 and the autocorrelation value is insignificant thereby resulting with a p-value larger than 0.05 suggesting no significant differences.

```{r, include=FALSE}
dwt(best.model)
```
```{r}
lag <- 1
auto <- -0.04748474
dw <- 2.09309
p_value <-  0.344

df_dwt <- data.frame(lag, auto, dw, p_value)

kable(df_dwt,
      col.names = c('Lag',
                    'Autocorrelation',
                    'D-W Statistic',
                    'p-value'),
      align = 'llll',
      caption = 'Durbin Watson Test')
```

# Discussion

```{r, include=FALSE}
model_all <- lm(df_res$`Sleep efficiency`~., df_res)
summary(model_all)
```

```{r, include=FALSE}
#Individual 
model1 <- lm(df_res$`Sleep efficiency`~df_res$`REM sleep percentage`, df_res)
model2 <- lm(df_res$`Sleep efficiency`~df_res$`Sleep duration`, df_res)
model3 <- lm(df_res$`Sleep efficiency`~df_res$`Deep sleep percentage`, df_res)
model4 <- lm(df_res$`Sleep efficiency`~df_res$Awakenings, df_res)
summary(model4)
```

It provides an overview of the statistical summary of the model with all corresponding response variables and a second model which respectively compare the independent variable to the dependent variable. Based on Table 5, Light Sleep Percentage had an error which was due to singularities therefore it appears as NA within the summary. At a glance, the sleep duration had a larger p-value associated with the T test (0.756) hence the variable does not have any significant impact on the model. Awakening 1 and REM Sleep percentage the was the second and third highest Pr(>|t|) 3.27e-10 and 6.94e-13 respectively. The remaining independent variable remain Pr(>|t|) value at < 2e-16 which shows strong evidence for significance differences. In addition, the initial model could explain 79% of the variability as it achieves an Adjusted R Square value of 0.7862. 
\newline

Upon comparing the statistical summary for the individual independent variable against the dependent variable. The sequence for least significance followed accordingly with the initial model (inclusion of all independent variables). The differences would be the Pr(>|t|) values which are 0.707, 0.208 and 3.55e-10 respectively. Moreover, the individual provides the adjusted R square value which provides detail regarding each independent variable variability towards the dependent variable. Based on table 6, the independent variable could only explain 0.22%, 1.52% & 39% of the variability for the dependent variable. Within the independent variable model and initial model, the awakening was subdivided into four groups where the number associated reflects the number of awakening during sleep. The variable was of categorical nature thus the split. 


```{r}
#Table 5: Multiple Linear Regression Analysis (Inclusion of Response Variable
model_all_no <- 1:11

model_all_variable <- c('Intercept',
                        'REM sleep percentage',
                        'Sleep duration',
                        'Deep sleep percentage',
                        'Light sleep percentage',
                        'Awakenings1',
                        'Awakenings2',
                        'Awakenings3',
                        'Awakenings4',
                        'Adjusted R-squared',
                        'p-value')

model_all_estimate <- c('0.3772824',
                        '0.0070902',
                        '0.0011312',
                        '0.0059880',
                        'NA',
                        '-0.0561390',
                        '-0.1314654',
                        '-0.1315943',
                        '-0.1240883',
                        '',
                        '')

model_all_pr <- c('< 2e-16',
                  '6.94e-13',
                  '0.756',
                  '< 2e-16',
                  'NA',
                  '3.27e-10',
                  '< 2e-16',
                  '< 2e-16',
                  '< 2e-16',
                  '',
                  '')

model_all_r <- c('',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '0.7862',
                 '')

model_all_p <- c('',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '',
                 '< 2.2e-16')

df_model_all <- data.frame(model_all_no,
                           model_all_variable,
                           model_all_estimate,
                           model_all_pr,
                           model_all_r,
                           model_all_p)

kable(df_model_all,
      align = 'llll',
      col.names = c('No',
                    'Variable',
                    'Estimate',
                    'Pr(>|t|)',
                    'Adjusted R Square',
                    'p-value'),
      caption = 'Multiple Linear Regression Analysis (Inclusion of Response Variable')
```

\newpage

```{r}
#Table 6: Simple Linear Regression Analysis (Individual Independent Variable
p_col <- c('Estimate',
           'Pr(>|t|)',
           'Adjusted R-squared',
           'p-value')

p_rem <- c('0.002533',
           '0.208',
           '0.001521',
           '0.2082')

p_duration <- c('-0.002944',
                '0.707',
                '-0.002222',
                '0.7067')

p_deepsleep <- c('0.0068769',
                 '<2e-16',
                 '0.6217',
                 '< 2.2e-16')

p_awake1 <- c('-0.09306',
              '3.55e-10',
              '0.3902  ',
              '< 2.2e-16')

p_awake2 <- c('-0.22064',
              '< 2e-16',
              '0.3902',
              '< 2.2e-16')

p_awake3 <- c('-0.21916',
              '< 2e-16',
              '0.3902',
              '< 2.2e-16')

p_awake4 <- c('-0.20832',
              '< 2e-16',
              '0.3902',
              '< 2.2e-16')


df_indi <- data.frame(p_col,
                      p_rem,
                      p_duration,
                      p_deepsleep,
                      p_awake1,
                      p_awake2,
                      p_awake3,
                      p_awake4)

kable(df_indi, 
      col.names = c('Independent Variable',
                             'REM Sleep %',
                             'Sleep Duration',
                             'Deep Sleep %',
                             'Awakening 1',
                             'Awakening 2',
                             'Awakening 3',
                             'Awakening 4'),
      align = 'llll',
      caption = 'Simple Linear Regression Analysis (Individual Independent Variable')

```

In the process of achieving the optimized model for the regression technique, Table 7 provides the result of the multiple regression analysis conducted on an iterative process. Since the project implements a backward feature selection, each independent variable is removed until the model with the best performance is identified. Table 7 allows the researcher to compare the result of the model based on comparison among the five models. On a glance, the p-value for all five models are identical therefore the p-value would not be the prime metric for comparison. Instead, the p-value associated with F-Statistic would be used as the prime metric. Model 3 had the highest value of 0.7555, it reflects that it was a strong model as it could perform well despite the removal of features. From the fourth model onwards, the value fell drastically, informing that there is a significant difference between the initial model due to feature removal. Moreover, the third model had the highest adjusted R square value of 0.7867 while model 1 and model 2 only achieved 0.7862 despite having additional features. It implies that the model was able to explain 1% more variability due to reduced complexity. Thus, the third model was selected due to these reasons.

```{r}
#Table 7: best model coefficient

m <- c('1','2','3','4','5')
a.r <- c('0.7862','0.7862', '0.7867', '0.6669','0.6217')
p.v <- c('2.2e-16','2.2e-16', '2.2e-16','2.2e-16','2.2e-16')
f.t <- c('0','0','0.7555','<2e-16','<2e-16')

#anova_models <- anova(model1, model2, model3, model4, model5)
#print(anova_models)



model_stats <- data.frame(m,a.r,f.t,p.v)
kable(model_stats,
      col.names = c('Model','Adjusted R-squared', 'Pr(>F)','p-value'),
      align = 'llll',
      caption = 'Simple Linear Regressional Model Summary')
```

# Limitation

As the dataset was collected by a group of university students in Morocco, it begs to question the quality of the dataset. Although the metadata does offer basic information regarding the data. It is unknown to the statistician about the procedure of data collecting as it is a secondary source. As the goal of the research was to commercialize the research into Apple.inc, there could be a potential gap in terms of data quality. In addition, the dataset only contains limited variables thus the research was incomplete as Apple Watch had additional features such as heart rate monitoring, blood oxygen monitoring etc which were not included within the studies due to the limited variable within the dataset. Future research could expand the dataset into a more diverse population and control the data quality to enhance the insight from the research. Moreover, the analytical approach mainly focused on linear relationships may possibly overlook complex, non-linear interactions between variables. 

# Conclusion

The researchers have successfully rejected the null hypothesis for the independent variable which were light sleep and sleep duration as the removal of the variable did not contribute any significant difference on the model. The null hypothesis for REM Sleep Percentage, Deep Sleep Percentage and Awakening failed to be rejected as these independent variables do significantly impact the models. Within the regression analysis, the research had implemented a simple and multiple linear regression model as the independent variable indicated an assumption that it could provide a linear relationship between the dependent and independent variable. The studies include an Exploratory Data Analysis to provide the reader the relationship from a visual perspective. Within the result, it is shown that these variable (REM Sleep Percentage, Deep Sleep Percentage & Awakening) was significantly important to the model as it explained 79% of the total variability while displaying the highest p-value associated with F-Statistic of value of 0.7555 which is indication of a strong model. In conclusion, the model does prove that the optimize model independent does contribute significantly different towards prediction sleep efficiency and further research could be implemented to commercialize the algorithms to provide better insight towards the user’s of apple.
\newpage

# References

1. Alfonsi, V., Gorgoni, M., Scarpelli, S., Zivi, P., Sdoia, S., Mari, E., Fraschetti, A., Ferlazzo, F., Giannini, A.M. and De Gennaro, L. (2021). COVID-19 lockdown and poor sleep quality: Not the whole story. Journal of Sleep Research. doi:https://doi.org/10.1111/jsr.13368.

2. Hockenberry, J. M., Curry, S. J., Fishman, P., Baker, T. B., Fraser, D., Cisler, R. A., … & Fiore, M. C. (2012). Healthcare costs around the time of smoking cessation. American Journal of Preventive Medicine, 42(6), 596-601. https://doi.org/10.1016/j.amepre.2012.02.019

3. Hofer, M. and Chen, F. (2020). The scent of a good night’s sleep: olfactory cues of a romantic partner improve sleep efficiency. Psychological Science, 31(4), 449-459. https://doi.org/10.1177/0956797620905615

4. Lee, S.W. (2022). Regression analysis for continuous independent variables in medical research: statistical standard and guideline of Life Cycle Committee. Life Cycle, 2. doi:https://doi.org/10.54724/lc.2022.e3.

5. Meltzer, L., Hiruma, L., Avis, K., Montgomery-Downs, H., & Valentin, J. (2015). Comparison of a commercial accelerometer with polysomnography and actigraphy in children and adolescents. Sleep, 38(8), 1323-1330. https://doi.org/10.5665/sleep.4918

6. Okamura, H., Mihara, K., Tsuda, A., Morisaki, T., Tanaka, Y., & Yamanaka, S. (2020). Subjective happiness is associated with objectively evaluated sleep efficiency and heart rate during sleep: an exploratory study using non-contact sheet sensors. Sustainability, 12(11), 4630. https://doi.org/10.3390/su12114630

7. Sharda, R., Delen, D., Turban, E. & King, D. (2017). Business Intelligence : a Managerial Approach, Global Edition. Harlow, United Kingdom: Pearson




